Parsing Postfix Log Files

### 2009-04-20 doc/logparser-background.tex
Corrections from proofreading for chapters 4 and 5.

That example also shows that it is trivial to supply a custom rejection
message.  When the result is the name of another restriction Postfix will
evaluate that restriction, allowing restrictions to be chosen based on the
client IP address, client hostname, HELO hostname, sender address, or
recipient address.  E.g.\ the administrator may require that all clients on
the local network have valid DNS entries, to prevent people sending mail
from unknown machines; one example of how to achieve this is:

@@

A policy server~\cite{policy-servers} is an external program consulted by
Postfix to determine the fate of an \acronym{SMTP} command.  The policy
server is given state information by Postfix (sample state information is
shown in \tableref{Example attributes sent to policy servers}) and returns
a result (reject, permit, dunno, or a rejection name) as described in
\sectionref{Mixing and matching Postfix restrictions}.  A policy server can
perform more complex checks than those provided by Postfix: a trivial
example is allowing addresses associated with the payroll system to send
mail on the third Tuesday after pay day only, to help prevent problems from
phishing mails using faked sender addresses.  E.g.\ a phishing mail might
claim that the payroll system had a disastrous disk failure, and until the
server is replaced all salary payments will have to be processed manually,
so please reply to this mail with your name, address, and bank account
details; the criminal can then use any details sent to him to help with
identity theft.

>>>>

That example also shows that it is trivial to supply a custom rejection
message.  When the result is the name of another restriction Postfix will
evaluate that restriction, allowing restrictions to be chosen based on the
client IP address, client hostname, HELO hostname, sender address, or
recipient address.  For example, the administrator may require that all
clients on the local network have valid DNS entries, to prevent people
sending mail from unknown machines; one example of how to achieve this is:

@@

A policy server~\cite{policy-servers} is an external program consulted by
Postfix to determine the fate of an \acronym{SMTP} command.  The policy
server is given state information by Postfix (sample state information is
shown in \tableref{Example attributes sent to policy servers}) and returns
a result (reject, permit, dunno, or a rejection name) as described in
\sectionref{Mixing and matching Postfix restrictions}.  A policy server can
perform more complex checks than those provided by Postfix: a trivial
example is allowing addresses associated with the payroll system to send
mail on the third Tuesday after pay day only, to help prevent problems from
phishing mails using faked sender addresses.  For example, a phishing mail
might claim that the payroll system had a disastrous disk failure, and
until the server is replaced all salary payments will have to be processed
manually, so please reply to this mail with your name, address, and bank
account details; the criminal can then use any details sent to him to help
with identity theft.

### 2009-04-20 doc/logparser-complications.tex
Corrections from proofreading for chapters 4 and 5.

It was initially expected that parsing Postfix log files would be a
relatively simple task, requiring a couple of months of work.  The author
had found Postfix log files useful when investigating problems reported by
users, and an examination of the log files gave the impression that they
would be straightforward to parse, process, and understand.  The large
variation in log lines was not apparent, because most log lines are
recognised by a small set of rules, as shown in \figureref{rule hits
graph}.  Most of the myriad complications and difficulties documented in
this section were discovered during \parsernames{} development, but the
first three complications were identified during the planning and design
phase of this project, influencing the architecture's design.

@@

    \item [Recreating Postfix behaviour] Three complications are caused by
        \parsername{} accurately recreating Postfix's behaviour.

    \item [Lack of information in the log files] Seven complications are
        caused by lack of information in the log files.  In general some
        extra logging by Postfix could remove these complications, and for
        one complication, extra logging in later versions of Postfix does
        remove it (\sectionref{identifying bounce notifications}); three of
        the complications would be removed if Postfix generated a log line
        whenever it discarded a mail.

    \item [Order log lines are found in log files] Five complications are
        caused by out of order log lines.  Postfix is not to blame for
        these, they are caused by process scheduling, inter-process
        communication delays, and are more likely to occur when the mail
        server is heavily loaded.

@@

@@

    \item Each action needs to know whether it deals with accepted mails,
        and should find the mail in the by\_queueid state tables, or deals
        with delivery attempts, and thus will find the connection in the
        by\_pid state tables.  Some actions, e.g.\
        \action{DELIVERY\_REJECTED}, need to query both tables.  Support
        functions fall into two groups: those that deal with state tables,
        and so need to know which one to operate on; the other group are
        passed a specific mail or connection to work with, and do not need
        to know about state tables at all.

@@

Multiple independent mails may be delivered across one connection: this
requires \parsername{} to clone the connection's data as soon as a mail is
accepted, so that subsequent mails will not overwrite each other's data.
This must be done every time a mail is accepted, as it is impossible to
tell in advance which connections will accept multiple mails.  Once a mail
has been accepted its log lines will not be correlated by \acronym{pid} any
more, its queueid will be used instead, as described in \sectionref{queueid
vs pid}.  If the original connection has any useful data (e.g.\ rejections)
it will be saved to the database when the client disconnects.  One unsolved
difficulty is distinguishing between different groups of rejections, e.g.\
when dealing with the following sequence:

@@

There should ideally be two separate entries in the database resulting from
the above sequence, but there will only be one.

@@

Mails sent to local addresses are not always delivered directly to a
mailbox: sometimes they are sent to and accepted for a local address, but
need to be delivered to one or more remote addresses due to aliasing.  When
this occurs, a child mail will be injected into the Postfix queue, but
without the explicit logging that mails injected by \daemon{smtpd} or
\daemon{postdrop} have.  Thus the source of the mail is not immediately
discernible from the log line in which the mail's queueid first appears:
from a strictly chronological reading of the log lines it usually appears
as if the child mail has been created without a source.  Subsequently the
parent mail will log the creation of the child mail, e.g.\ parent mail
\texttt{3FF7C4317} creates child mail \texttt{56F5B43FD}\@:

@@

Unfortunately, although all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition, the first log line belonging to the child mail (the example log
line above belongs to the parent mail) is logged by either \daemon{qmgr} or
\daemon{cleanup}, so the order also depends on how soon they process the
new mail.

Because of the uncertain order that log lines can appear in, \parsername{}
cannot complain when it encounters a log line from either \daemon{qmgr} or
\daemon{cleanup} for a previously unseen mail; instead it must flag the
mail as coming from an unknown origin, and subsequently clear that flag if
the origin of the mail becomes clear.  Obviously the parser could omit
checking where mails originate from, but requiring an explicit source helps
to expose bugs in the parser; such checks helped to identify the
complications described in \sectionref{discarding cleanup log lines} and
\sectionref{pickup logging after cleanup}.

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered, and entirely finished with,
\textit{before\/} the parent mail logs its creation!  Thus, mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is processed, and \parsername{} cannot warn the user;
instead they must be marked as ready for entry and subsequently entered
once their origin has been identified.  The crux of this complication is
that re-injected mails appear in the log files without explicit logging
indicating their source.  Unfortunately, it is not possible to distinguish
between child mails waiting to be tracked and other mails whose origin is
unknown (except for bounce notifications, as described in
\sectionref{identifying bounce notifications}); postponing the entry into
the database of mails whose origin is unknown can result in those mails
remaining indefinitely in the state tables, causing a queueid clash if the
queueid is reused, resource usage for the period they remain in the state
tables, and, most importantly, \parsernames{} understanding of such mails
is incorrect.  This problem has only been observed when a state table entry
is missing its initial log line, usually because it is in an earlier log
file; this specific instance is not a serious problem, because
\parsername{} can not be expected to fully understand and reconstruct a
mail when it is lacking some of that mail's log lines.

@@

    \item If a mail is tagged with the unknown origin flag, it is assumed
        to be a child mail whose parent has not yet been identified.  The
        mail is tagged as ready to be entered in the database, but entry is
        postponed until the parent is identified.  The child mail should
        not have any subsequent log lines: only its parent will refer to
        it.

@@

Postfix 2.2.x (and presumably previous versions) does not generate a log
line when it generates a bounce notification; suddenly there will be log
entries for a mail that lacks an obvious source.  Similarities exist to the
problem of identifying re-injected mails discussed in \sectionref{tracking
re-injected mail}, but unlike the solution described therein bounce
notifications do not eventually have a log line that identifies their
source.  Heuristics must be used to identify bounce notifications:

@@

    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
        messages associated with the mail, indicating it was generated
        internally by Postfix, not accepted via \acronym{SMTP} or submitted
        locally by \daemon{postdrop}.

@@

    \item The queueid in the message-id must be the same as the queueid of
        the mail: this is what distinguishes a new bounce notification from
        a bounce notification that is being re-injected as a result of
        aliasing.  For the latter, the message-id will be unchanged from
        the original bounce notification, and so even if it happens to be
        in the correct format, e.g.\ if it was generated by Postfix on this
        or another server, it will not correspond with the queueid of the
        mail.

@@

If a mail is misclassified as a bounce message it will almost certainly
have been generated internally by Postfix; arguably misclassification of
this kind is a benefit rather than a drawback, as other mails generated
internally by Postfix will be handled correctly.  Postfix 2.3 and
subsequent versions log the creation of a bounce message.

@@

Postfix does not log a message making the client's behaviour clear, so
again heuristics are required to identify when a delivery attempt is
aborted in this way.  A list of all mails accepted during a connection is
saved in the connection's state table entry, and the accepted mails are
examined when the disconnection action is invoked.  Each mail is checked
for the following:

@@

If all the checks above are successful then the mail is assumed to be an
aborted delivery attempt and is discarded.  There will be no further
entries logged for such mails, so without identifying and discarding them
they accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and IP address, but the database schema
requires many more fields be populated (see \sectionref{connections table}
and \sectionref{results table}).  These heuristics are quite restrictive,
and it appears to have little scope for false positives; any false
positives will cause a warning when the next log line for that mail is
parsed.  False negatives are less likely to be detected: there may be
queueid clashes (and warnings) if mails remain in the state tables after
they should have been removed, otherwise the only way to detect false
negatives is to examine the state tables after each parsing run.

@@

If both checks are successful then the mail is assumed to have been
discarded when the client disconnected.  There will be no further log lines
for such mails, so without identifying and entering them in the database
immediately they accumulate in the state table and will cause clashes if
the queueid is reused.

@@

    \item If one or more recipients have been accepted, Postfix will have
        allocated a queueid for the incoming mail, and there may be a mail
        in the state tables that needs to be discarded.  The timeout may
        have interrupted transfer of an accepted delivery attempt, or it
        may have occurred after a mail delivery attempt was rejected.  If a
        mail needs to be discarded, the following checks will all pass:

@@

        If all the checks are passed the mail will be discarded from the
        state table, and will not be entered in the database.  If one or
        more checks are not passed it is assumed that the timeout happened
        after a rejected delivery attempt.  This assumption is not
        necessarily correct, because Postfix may have accepted an earlier
        recipient and rejected a later one, so the timeout applies to the
        partially accepted mail, which should be discarded; this has not
        been a problem in practice.  Processing timeouts is further
        complicated by the presence of out of order \daemon{cleanup} log
        lines: see \sectionref{discarding cleanup log lines} for details.

@@

This complication is dealt with by the \action{TIMEOUT} action, with help
from the \action{CLONE} action.

@@

During the \action{TIMEOUT} action, if the \daemon{cleanup} log line is
still pending, the action updates the \texttt{timeout\_queueids} state
table, adding the queueid and the timestamp from the log line.  When the
next \daemon{cleanup} log line is parsed for that queueid, the list will be
checked and if the log line meets the following requirements it will be
deemed part of the mail where the timeout occurred and discarded.

@@

    \item The queueid must not have been reused yet, i.e.\ it does not have
        a mail in the state tables.

@@

The \daemon{cleanup} log line must meet the criteria above for it to be
discarded because some, but not all, connections where a timeout occurs
will have an associated \daemon{cleanup} log line; if the
\action{CLEANUP\_PROCESSING} action blindly discarded the next
\daemon{cleanup} log line after a timeout it would sometimes be mistaken.
When the next \daemon{pickup} log line containing that queueid is processed
the queueid will be removed from the cache of timeout queueids, regardless
of whether it meets the criteria above.

@@

When mail is submitted locally, \daemon{pickup} accepts the new mail and
generates a log line.  Occasionally this log line will occur later in the
log file than the \daemon{cleanup} log line, so the \action{PICKUP} action
will find that a state table entry already exists for that queueid.
Normally when this happens a warning is generated by the \action{PICKUP}
action, but if the following conditions are met it is assumed that the log
lines were out of order:

@@

Occasionally a \daemon{smtpd} will just stop logging, without an
immediately obvious reason.  After poring over log files for some time
several reasons have been found for this rare event:

@@

    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
        for some reason, by the administrator, or by the OS), so its active
        connection must be cleaned up, entered in the database if it has
        enough data, and deleted from the connections state table.

@@

Occasionally a log file will have out of order log lines that cannot be
dealt with by the techniques described in \sectionref{tracking re-injected
mail}, \sectionref{discarding cleanup log lines}, or \sectionref{pickup
logging after cleanup}.  In the \numberOFlogFILES{} log files used for
testing this problem occurs only five times in 60,721,709 log lines, but
for parser correctness it must be dealt with.  All five occurrences have
the same characteristics: the \daemon{local} log line showing delivery to a
local mailbox comes after the \daemon{qmgr} log line showing removal of the
mail from the queue because delivery is complete.  This causes problems:
the data in the state tables for the mail is not complete, so it cannot be
entered into the database; a new mail is created when the \daemon{local}
log line is processed and remains in the state tables; four warnings are
issued per pair of out of order log lines.

The \action{COMMIT} action examines the list of programs that have produced
log lines for each mail, checking the list against a table of known-good
Postfix component combinations.  If the mail's combination is found in the
table the mail can be entered in the database; if the combination is not
found entry must be postponed and the mail flagged for later entry.  The
\action{MAIL\_DELIVERED} action checks for that flag; if the additional log
lines have caused the mail to reach a valid combination then entry in the
database will proceed, otherwise it must be postponed once more.

The list of valid combinations is explained below.  Every mail will
additionally have log lines from \daemon{cleanup} and \daemon{qmgr}; any
mail may also have a log line from \daemon{bounce}, \daemon{postsuper}, or
both.

@@

    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
        client, then remotely delivered (typically relaying mail for
        clients on the local network to addresses outside the local
        network).

    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
        client, then deleted by the administrator before any delivery
        attempt was made.  Notice that \daemon{postsuper} is required, not
        optional, for this combination.

@@

The aborted delivery attempts described in \sectionref{aborted delivery
attempts} occur frequently, but those described in this section only occur
four times in the \numberOFlogFILES{} log files used for testing.  The
symptoms are the same as in \sectionref{aborted delivery attempts}, except
that the mail has a \daemon{cleanup} log line; nothing can be found in the
log file to explain why this mail does not have any further log lines.  The
only way to detect these mails is to periodically scan all mails in the
state tables, deleting any mails with the following characteristics:

@@

Postfix logs the recipient address when delivery of a mail is attempted, so
if delivery has yet to be attempted \parsername{} cannot determine the
recipient address or addresses.  This is a problem when mail is arriving
faster than Postfix can attempt delivery, and the administrator deletes
some of the mail before Postfix has had a chance to try to deliver it.  The
recipient address will not have been logged, so a dummy recipient address
needs to be added, because every mail is required by the database schema
(\sectionref{results table}) to have at least one recipient.  Typically
when this complication occurs the log file will show many instances of it,
closely grouped.

This lack of information cannot easily be overcome: it is trivial to log a
warning for every accepted recipient, but Postfix will not yet have
allocated a queueid for the mail when the warning for the first recipient
is logged, so the warning will be associated with the connection rather
than the accepted mail.  A queue file and queueid will be allocated after
Postfix accepts the MAIL FROM command if
\texttt{smtpd\_delay\_open\_until\_valid\_rcpt} is set to ``no'', but that
setting will cause disk IO for every delivery attempt, instead of just for
delivery attempts where recipients are accepted, and consequently a drastic
reduction in the performance of the mail server.

@@

The administrator can delete mails using \daemon{postsuper}; occasionally
mails that are in the process of being delivered will be deleted.  This
results in the log lines from the delivery agent (\daemon{local},
\daemon{virtual} or \daemon{smtp}) appearing in the log file
\textit{after\/} the mail has been removed from the state tables and saved
in the database.  The \action{DELETE} action adds deleted mails to a cache,
which is checked by the \action{SAVE\_DATA} action, and the current log
line discarded if the following conditions are met:

@@

This section has described the complications encountered while implementing
\parsername{}.  Five of the fifteen complications are caused by log lines
appearing in an unexpected or abnormal order in log files, often because of
heavy system load: these complications typically caused warnings and a new
state table entry.  Three of the fifteen complications are required to
correctly recreate Postfix's behaviour.  The remaining seven complications
are caused by deficiencies in Postfix's logging, and some could easily be
resolved by adding additional logging to Postfix; these seven are dealt
with by applying heuristics to specific log lines or mails, sometimes in
conjunction with information cached when mails are removed from the state
tables.

>>>>

It was initially expected that parsing Postfix log files would be a
relatively simple task, requiring a couple of months of work.  The author
had found Postfix log files useful when investigating problems reported by
users, and an examination of several log files gave the impression that
they would be straightforward to parse, process, and understand.  The large
variation in log lines was not apparent, because most log lines are
recognised by a small set of rules, as shown in \figureref{rule hits
graph}.  Most of the myriad complications and difficulties documented in
this section were discovered during \parsernames{} development, but the
first three complications were identified during the planning and design
phase of this project, influencing the architecture's design.

@@

    \item [Recreating Postfix behaviour] Three complications needed to be
        solved for \parsername{} to accurately recreate Postfix's
        behaviour.

    \item [Lack of information in the log files] Seven complications are
        caused by lack of information in the log files.  In general, some
        extra logging by Postfix could remove these complications, and for
        one complication, extra logging in later versions of Postfix does
        remove it (\sectionref{identifying bounce notifications}); three of
        the complications would be removed if Postfix generated a log line
        whenever it discarded a mail.

    \item [The order that log lines are found in log files] Five
        complications are caused by out of order log lines.  Postfix is not
        to blame for these: they are caused by process scheduling,
        inter-process communication delays, and are more likely to occur
        when the mail server is heavily loaded.

@@

    \squeezeitems{}

@@

    \item Each action needs to know whether it deals with accepted mails,
        and should find the mail in the by\_queueid state tables, or deals
        with delivery attempts, and thus will find the connection in the
        by\_pid state tables.  Some actions, e.g.\
        \action{DELIVERY\_REJECTED}, need to query both tables.  Support
        functions fall into two groups: those that deal with state tables,
        and so need to know which one to operate on; the functions in the
        other group are passed a specific mail or connection to work with,
        and do not need to know about state tables at all.

@@

Multiple independent mails may be delivered across one connection: this
requires \parsername{} to clone the connection's data as soon as a mail is
accepted, so that subsequent mails will not overwrite each other's data.
This must be done every time a mail is accepted, because it is impossible
to tell in advance which connections will accept multiple mails.  Once a
mail has been accepted its log lines will not be correlated by
\acronym{pid} any more: its queueid will be used instead, as described in
\sectionref{queueid vs pid}.  If the original connection has any useful
data (e.g.\ rejections) it will be saved to the database when the client
disconnects.  One unsolved difficulty is distinguishing between different
groups of rejections, e.g.\ when dealing with the following sequence:

@@

Ideally, the sequence above would result in two separate entries in the
connections table, but only one will be created.

@@

Mails sent to local addresses are not always delivered directly to a
mailbox: sometimes they are sent to and accepted for a local address, but
need to be delivered to one or more remote addresses because of aliasing.
When this occurs, a child mail will be injected into the Postfix queue, but
without the explicit logging that mails injected by \daemon{smtpd} or
\daemon{postdrop} have.  Thus the source of the mail is not immediately
discernible from the log line in which the mail's queueid first appears:
from a strictly chronological reading of the log lines it usually appears
as if the child mail has been created without a source.  Subsequently, the
parent mail will log the creation of the child mail, e.g.\ parent mail
\texttt{3FF7C4317} creates child mail \texttt{56F5B43FD}\@:

@@

The sample log line above would be processed by the \action{TRACK} action,
which creates the child mail if it does not exist in the state tables,
links it to the parent mail, and checks that the child is not being tracked
for a second time.

Unfortunately, although all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved in a log file is subject to the vagaries of process
scheduling.  In addition, the first log line belonging to the child mail
(the example log line above belongs to the parent mail) is logged by either
\daemon{qmgr} or \daemon{cleanup}, so the order also depends on how soon
they process the new mail.

Because of the uncertain order that log lines can appear in, \parsername{}
cannot complain when it encounters a log line from either \daemon{qmgr} or
\daemon{cleanup} for a mail that does not exist in the state tables;
instead it must create the state table entry and flag the mail as coming
from an unknown origin.  Subsequently invoked actions will clear that flag
if the origin of the mail becomes clear.  Obviously the parser could omit
checking where mails originate from, but requiring an explicit source helps
to expose bugs in the parser; such checks helped to identify the
complications described in \sectionref{discarding cleanup log lines} and
\sectionref{pickup logging after cleanup}.

Process scheduling can have a still more confusing effect: the child mail
will often be created, delivered, and entirely finished with,
\textit{before\/} the parent mail logs its creation!  Thus, mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is processed, and \parsername{} cannot warn the user;
instead they must be marked as ready for entry and subsequently entered
once their origin has been identified.  Unfortunately, it is not possible
to distinguish between child mails waiting to be tracked and other mails
whose origin is unknown (except for bounce notifications, as described in
\sectionref{identifying bounce notifications}); postponing the entry into
the database of mails whose origin is unknown can result in those mails
remaining indefinitely in the state tables, causing a queueid clash if the
queueid is reused, resource usage for the period they remain in the state
tables, and, most importantly, \parsernames{} understanding of such mails
is incorrect.  This problem has only been observed when a state table entry
is missing its initial log line, usually because it is in an earlier log
file; this specific instance is not a serious problem, because
\parsername{} can not be expected to fully understand and reconstruct a
mail when it is lacking some of that mail's log lines.

@@

    \item If a mail is tagged with the unknown origin flag, it is assumed
        to be a child mail whose parent has not yet been identified.  The
        mail is tagged as ready to be entered in the database, but entry is
        postponed until the parent is identified.  The child mail will not
        have any subsequent log lines: only its parent will refer to it.

@@

Postfix 2.2.x (and presumably previous versions) does not generate a log
line when it generates a bounce notification; the lof file will have log
lines for a mail that lacks an obvious source.  Similarities exist to the
problem of identifying re-injected mails discussed in \sectionref{tracking
re-injected mail}, but unlike the solution described therein bounce
notifications do not eventually have a log line that identifies their
source.  Heuristics must be used to identify bounce notifications:

@@

    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
        messages associated with the mail, indicating that it was generated
        internally by Postfix, rather than accepted via \acronym{SMTP} or
        submitted locally by \daemon{postdrop}.

@@

    \item The queueid embedded in the message-id must be the same as the
        queueid of the mail: this is what distinguishes a new bounce
        notification from a bounce notification that is being re-injected
        as a result of aliasing.  For the latter, the message-id will be
        unchanged from the original bounce notification, and so even if it
        happens to be in the correct format, i.e.\ if it was generated by
        Postfix on this or another server, it will not equal the queueid of
        the mail.

@@

If a mail is misclassified as a bounce message it will almost certainly
have been generated internally by Postfix; arguably misclassification of
this kind is a benefit rather than a drawback, because other mails
generated internally by Postfix will be handled correctly.  Postfix 2.3 and
subsequent versions log the creation of a bounce message.

@@

Postfix does not log a message making the client's behaviour clear, so
again heuristics are required to identify when a delivery attempt is
aborted in this way.  A list of all mails accepted during a connection is
saved in the connection's state table entry, and the accepted mails are
examined when the disconnection action is invoked.  Each accepted mail is
checked for the following:

@@

If all the checks above are successful then the mail is assumed to be an
aborted delivery attempt and is removed form the state tables.  There will
be no further entries logged for such mails, so without identifying and
discarding them they accumulate in the state table and will cause clashes
if the queueid is reused.  The mail cannot be entered in the database
because the only data available is the client hostname and IP address, but
the database schema requires many more fields be populated (see
\sectionref{connections table} and \sectionref{results table}).  These
heuristics are quite restrictive, and appear to have little scope for false
positives; any false positives would cause a warning when the next log line
for that mail is parsed.  False negatives are less likely to be detected:
there may be queueid clashes (and warnings) if mails remain in the state
tables after they should have been removed, otherwise the only way to
detect false negatives is to examine the state tables after each parsing
run.

@@

If both checks are successful then the mail is assumed to have been
discarded by Postfix when the client disconnected; \parsername{} will
remove it from the state tables.  There will be no further log lines for
such mails, so if \parsername{} does not deal with them immediately they
accumulate in the state table and will cause clashes if the queueid is
reused.

@@

    \item If one or more recipients have been accepted, Postfix will have
        allocated a queueid for the incoming mail, and there may be a mail
        in the state tables that needs to be discarded by \parsername{}.  The timeout may
        have interrupted transfer of an accepted delivery attempt, or it
        may have occurred after a mail delivery attempt was rejected.  If a
        mail needs to be discarded, the following checks will all pass:

@@

        If all the checks pass the mail will be discarded from the state
        tables, and will not be entered in the database.  If one or more
        checks do not pass it is assumed that the timeout happened after a
        rejected delivery attempt.  This assumption is not necessarily
        correct, because Postfix could have accepted an earlier recipient,
        rejected a later one, and continued to accept delivery of the mail
        for the first recipient.  Tn that case the timeout applies to the
        partially accepted mail, which should be discarded; however, this
        has not occurred in practice.  Processing timeouts is further
        complicated by the presence of out of order \daemon{cleanup} log
        lines: see \sectionref{discarding cleanup log lines} for details.

@@

This complication is dealt with by the \action{TIMEOUT} action, with help
from the \action{CLONE} action.  When a client tries to delivery a mail
larger than the sever will accept, the \action{MAIL\_TOO\_LARGE} action
will perform the same processing as the \action{TIMEOUT} action.

@@

During the \action{TIMEOUT} action, if the \daemon{cleanup} log line is
still pending, the action updates the \texttt{timeout\_queueids} state
table, adding the queueid and the timestamp from the log line.  The
following checks will be performed for each \daemon{cleanup} log line
that is processed:

@@

    \item Does the \texttt{timeout\_queueids} state table have an entry for
        the queueid in the log line?  If there is an entry, it will be
        removed, regardless of whether the remaining criteria are
        satisfied.

    \item Has the queueid been reused yet, i.e.\ does it have an entry in
        the state tables?

@@

The \daemon{cleanup} log line must meet the criteria above for it to be
discarded because some, but not all, connections where a timeout occurs
will have an associated \daemon{cleanup} log line; if the
\action{CLEANUP\_PROCESSING} action blindly discarded the next
\daemon{cleanup} log line after a timeout it would sometimes be mistaken.

@@

When mail is submitted locally, \daemon{pickup} processes the new mail and
generates a log line.  Occasionally this log line will occur later in the
log file than the \daemon{cleanup} log line, so the \action{PICKUP} action
will find that a state table entry already exists for that queueid.
Normally when this happens a warning is generated by the \action{PICKUP}
action, but if the following conditions are met it is assumed that the log
lines were out of order:

@@

Occasionally a \daemon{smtpd} will stop logging without an immediately
obvious reason.  After poring over log files for some time, several reasons
have been found for this rare event:

@@

    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by
        Postfix, by the administrator, or by the OS), so its active
        connection must be cleaned up, entered in the database if it has
        enough data, and deleted from the connections state table.

@@

Occasionally a log file will have out of order log lines that cannot be
dealt with by the techniques described in \sectionref{tracking re-injected
mail}, \sectionref{discarding cleanup log lines}, or \sectionref{pickup
logging after cleanup}.  In the \numberOFlogFILES{} log files used for
testing this problem occurs only five times in 60,721,709 log lines.  All
five occurrences have the same characteristics: the \daemon{local} log line
showing delivery to a local mailbox comes after the \daemon{qmgr} log line
showing removal of the mail from the queue because delivery is complete.
This causes problems: the data in the state tables for the mail is not
complete, so it cannot be entered into the database; a new mail is created
when the \daemon{local} log line is processed and remains in the state
tables; four warnings are issued per pair of out of order log lines.

The \action{COMMIT} action examines the list of programs that have produced
log lines for each mail, checking the list against a table of known-good
Postfix component combinations.  If the mail's combination is found in the
table it can be entered in the database; if the combination is not found
entry must be postponed and the mail flagged for later entry.  The
\action{MAIL\_DELIVERED} action checks for that flag; if the additional log
lines have caused the mail to reach a valid combination then its entry in
the database will proceed, otherwise it must be postponed once more.

The valid combinations are explained below.  In addition to the components
shown in each combination, every mail will have log lines from both
\daemon{cleanup} and \daemon{qmgr}, and any mail may also have a log line
from \daemon{bounce}, \daemon{postsuper}, or both.

@@

    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
        client, then remotely delivered.  Typically this is a mail server
        relaying mail for clients on the local network to addresses outside
        the local network.

    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
        client, then deleted by the administrator before any delivery
        attempt was made.  Note that \daemon{postsuper} is required, not
        optional, for this combination.

@@

The aborted delivery attempts described in \sectionref{aborted delivery
attempts} occur frequently, but those described in this section only occur
four times in the \numberOFlogFILES{} log files used for testing.  The
symptoms are the same as in \sectionref{aborted delivery attempts}, except
that the mail \textit{has\/} a \daemon{cleanup} log line; nothing can be
found in the log files to explain why this mail does not have any further
log lines.  The only way to detect these mails is to periodically scan all
mails in the state tables, deleting any mails with the following
characteristics:

@@

Postfix logs the recipient address when delivery of a mail is attempted, so
if delivery has yet to be attempted \parsername{} cannot determine the
recipient address or addresses.  This is a problem when mail is arriving
faster than Postfix can attempt delivery, and the administrator deletes
some of the mail before Postfix has had a chance to attempt delivery.  The
recipient address will not have been logged yet, and the deleted mails will
not have any more log lines for it to be logged in.  A dummy recipient
address needs to be added, because every mail is required by the database
schema (\sectionref{results table}) to have at least one recipient.
Typically when this complication occurs the log file will show many
instances of it, closely grouped.

This lack of information cannot easily be overcome: it is trivial to log a
warning for every accepted recipient, but Postfix will not yet have
allocated a queueid for the mail when the warning for the first recipient
is logged, so the warning will be associated with the connection rather
than the accepted mail.  A queueid will be allocated after Postfix accepts
the MAIL FROM command if \texttt{smtpd\_delay\_open\_until\_valid\_rcpt} is
set to ``no'', but that setting will cause disk IO for every delivery
attempt, instead of just for delivery attempts where recipients are
accepted, and consequently a drastic reduction in the performance of the
mail server.

@@

The administrator can delete mails using \daemon{postsuper}; occasionally
mails that are in the process of being delivered will be deleted.  This
results in the log lines from the delivery agent (\daemon{local},
\daemon{smtp}, or \daemon{virtual}) appearing in the log file
\textit{after\/} the mail has been removed from the state tables and saved
in the database.  The \action{DELETE} action adds deleted mails to a cache,
which is checked by the \action{SAVE\_DATA} action, and the current log
line discarded if the following conditions are met:

@@

This section has described the complications encountered while implementing
\parsername{}.  Five of the fifteen complications are caused by log lines
appearing in an unexpected or abnormal order in log files, often because of
heavy system load: these complications typically caused warnings and a new
state table entry.  Three of the fifteen complications must be dealt with
to correctly recreate Postfix's behaviour.  The remaining seven
complications are caused by deficiencies in Postfix's logging, and some
could easily be resolved by adding additional logging to Postfix; these
seven are dealt with by applying heuristics to specific log lines or mails,
sometimes in conjunction with information cached when mails are removed
from the state tables.
